%-----------------------------------------------------------------------------
%
%               Template for sigplanconf LaTeX Class
%
% Name:         sigplanconf-template.tex
%
% Purpose:      A template for sigplanconf.cls, which is a LaTeX 2e class
%               file for SIGPLAN conference proceedings.
%
% Guide:        Refer to "Author's Guide to the ACM SIGPLAN Class,"
%               sigplanconf-guide.pdf
%
% Author:       Paul C. Anagnostopoulos
%               Windfall Software
%               978 371-2316
%               paul@windfall.com
%
% Created:      15 February 2005
%
%-----------------------------------------------------------------------------


%\documentclass[preprint]{sigplanconf}
\documentclass{sigplanconf}

% The following \documentclass options may be useful:

% preprint      Remove this option only once the paper is in final form.
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% numbers       To obtain numeric citation style instead of author/year.


%%%%%%%%%%%%%%%%%%%%%%%%%
% PACKAGES
%%%%%%%%%%%%%%%%%%%%%%%%%

\usepackage{listings}
\usepackage{multicol}
%\usepackage{stmaryrd}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{proof}
%\usepackage{lscape}
\usepackage{url}


%\mathchardef\mhyphen="2D % Define a "math hyphen"


\input{formatting}


\newcommand{\cL}{{\cal L}}
\newcommand{\foldn}{\ensuremath{FO\lambda^{\Delta I\!\!N}}}

\begin{document}

\special{papersize=8.5in,11in}
\setlength{\pdfpageheight}{\paperheight}
\setlength{\pdfpagewidth}{\paperwidth}

\conferenceinfo{LFMTP '16}{June 23, 2016, Porto, Portugal}
\copyrightyear{2016}
%\copyrightdata{978-1-nnnn-nnnn-n/yy/mm}
\copyrightdata{978-1-4503-4777-8/16/06}
\copyrightdoi{http://dx.doi.org/10.1145/2966268.2966271}

% Uncomment the publication rights you want to use.
\publicationrights{transferred}
%\publicationrights{licensed}     % this is the default
%\publicationrights{author-pays}

%\titlebanner{banner above paper title}        % These are ignored unless
%\preprintfooter{short description of paper}   % 'preprint' option specified.

\title{The Logic of Hereditary Harrop Formulas as a Specification
  Logic for Hybrid}
%\title{Two-Level Reasoning in Hybrid with Higher-Order Logic}
%\subtitle{}

\authorinfo{Chelsea Battell}
           {Department of Mathematics and Statistics\\
            University of Ottawa, Canada}
           {cbattell@uottawa.ca}
\authorinfo{Amy Felty}
           {School of Electrical Engineering and Computer Science\\
            and Department of Mathematics and Statistics\\
            University of Ottawa, Canada}
           {afelty@uottawa.ca}

\maketitle

\begin{abstract}
  Hybrid is a logical framework that supports the use of higher-order
  abstract syntax (HOAS) in representing formal systems or ``object
  logics'' (OLs).  It is implemented in Coq and follows a two-level
  approach, where a specification logic (SL) is implemented as an
  inductive type and used to concisely and elegantly encode the
  inference rules of the formal systems of interest.  In this paper,
  we develop a new higher-order specification logic for Hybrid.  By
  increasing the expressive power of the SL beyond what was considered
  previously, we increase the flexibility of encoding OLs and thus
  extend the class of formal systems for which we can reason about
  efficiently.  We focus on formalizing the meta-theory of the SL.
% new:
  We develop an abstract way in which to present an important class of
  meta-theorems.  This class includes
%
  properties such as weakening, contraction, exchange, and the
  admissibility of the cut rule.  The cut admissibility theorem
  establishes consistency and also provides justification for
  substituting a formula for an assumption in a context of
  assumptions.  It can greatly simplify reasoning about OLs in systems
  that provide HOAS.
%  We discuss the proof of this theorem in detail, illustrating the
%  challenges of its formalization.
% new:
  We present the abstraction and show how it is used to prove all of
  these theorems.
\end{abstract}

%\category{CR-number}{subcategory}{third-level}
\category{F.3.1}{Logics and Meanings of Programs}{Specifying and
Verifying and Reasoning about Programs}
\category{F.4.1}{Mathematical Logic and Formal Languages}{Mathematical
Logic}

% general terms are not compulsory anymore,
% you may leave them out
%\terms
%term1, term2

\keywords
logical frameworks, higher-order abstract syntax, cut admissibility,
structural rules, interactive theorem proving, Coq

\input{intro}

\input{hybrid}                                                                 

\input{sl}

\input{structind}

\input{sl_generalized}

\input{gsl_ind}

\input{structrules}

\input{cutadmiss}

\input{conclusion}

%\appendix
%\section{Appendix Title}
%
%This is the text of the appendix, if you need one.

\acks

\input{acks}

% We recommend abbrvnat bibliography style.

\bibliographystyle{abbrvnat}
\bibliography{lfmtp16}   

% The bibliography should be embedded for final submission.

\end{document}
